---
title: "Assignment 5: Data Visualization"
author: "Siyu Dong"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r Initialize project}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

getwd()

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
NiwotRidgeLitter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
NiwotRidgeLitter$collectDate <- as.Date(NiwotRidgeLitter$collectDate, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
```

## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r Build plot theme}
#3
library(ggplot2)

ivy_theme <-  theme_classic()+
  theme(plot.title = element_text(color = "darkblue", size = 16, face = "bold", hjust = 0.5),
        axis.title.x = element_text(color = "darkblue", size = 12),
        axis.title.y = element_text(color = "darkblue", size = 12),
        legend.position = "right",
        legend.title = element_text(color = "darkblue", size = 12),
        legend.text = element_text(color = "darkblue", size = 10))

# Set custom theme as default theme
theme_set(ivy_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r Plot total P vs PO4}
#4 
# First preliminarily plot  total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes to check if there are any extreme values.
ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug)) +
  geom_point()

# Most of the observations have the po4 values within 50. The only extreme observation is around (po4=375, tp_ug=75), which influces the limit of x axis setting greatly. Therefore, here I set xlim to (0,50) and remain the same for tp_ug.
ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug)) +
  geom_point() +
  xlim(0, 50) +
  geom_smooth(method = lm)

# Add an abline on the basis of previous plot and adjust aesthetics
P_PointLine<- ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug)) +
  geom_point() +
  xlim(0, 50) +
  geom_smooth(method = lm) +
  geom_abline(aes(slope = 2.7, intercept = 0), alpha = 0.3) +
  labs(title = "Total Phosphorus by Phosphate") +
  ivy_theme

print(P_PointLine)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
NTL_LTER_BoxA <- ggplot(PeterPaul.chem.nutrients, aes(x = factor(month), y = temperature_C, color = lakename))+
  geom_boxplot() + 
  labs(title = "Lake Temperature by Month", x = "Month", y = "Temperature(°C)") +
  ivy_theme

NTL_LTER_BoxB <- ggplot(PeterPaul.chem.nutrients, aes(x = factor(month), y = tp_ug, color = lakename))+
  geom_boxplot() + 
  labs(title = "Lake Phosphorus by Month", x = "Month", y = "Total Phosphorus") +
  ivy_theme

NTL_LTER_BoxC <- ggplot(PeterPaul.chem.nutrients, aes(x = factor(month), y = tp_ug, color = lakename))+
  geom_boxplot() + 
  labs(title = "Lake Nitrogen by Month", x = "Month", y = "Total Nitrogen") +
  ivy_theme

library(cowplot)

# Combine the individual plots
combined_plot <- plot_grid(
  NTL_LTER_BoxA + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()),
  NTL_LTER_BoxB + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank()),
  NTL_LTER_BoxC + theme(axis.title.x = element_blank(), axis.title.y = element_blank()),
  align = "hv", labels = c("A", "B", "C")
)

# Print the combined plot
print(combined_plot)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: