---
title: "Assignment 10: Data Scraping"
author: "Siyu Dong"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE, error = FALSE, warning = FALSE, echo = TRUEE}
#1 
library(tidyverse)
library(rvest)
library(ggplot2)

getwd()

mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
Durham_LWSP_2022web <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
Durham_LWSP_2022web
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Durham_LWSP_WaterSystem <- Durham_LWSP_2022web %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
Durham_LWSP_WaterSystem

Durham_LWSP_PWSID <- Durham_LWSP_2022web %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
Durham_LWSP_PWSID

Durham_LWSP_Ownership <- Durham_LWSP_2022web %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Durham_LWSP_Ownership

Durham_LWSP_MaxWithdraws <- Durham_LWSP_2022web %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
Durham_LWSP_MaxWithdraws
class(Durham_LWSP_MaxWithdraws)
```

4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
#Since withdraws were not scraped in chronological order, the corresponding month vector will be scraped here.
LWSP_Month <- Durham_LWSP_2022web %>% 
  html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% 
  html_text()
LWSP_Month

months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

LWSP_Month_numeric <- as.numeric(match(LWSP_Month, months))

#Create a dataframe. 
Durham_LWSP_df <- data.frame("Year" = rep(2022, 12))

#Modify the dataframe
Durham_LWSP_df <- Durham_LWSP_df %>% 
  mutate(WaterSystem_Name = !!Durham_LWSP_WaterSystem,
         PWSID = !!Durham_LWSP_PWSID,
         Ownership = !!Durham_LWSP_Ownership,
         MaxDayUse = as.numeric(Durham_LWSP_MaxWithdraws),
         Date = my(paste(LWSP_Month_numeric,"-",Year)))

#5 
ggplot(Durham_LWSP_df, aes(x = Date, y = MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Maximum Daily Withdrawals for", Durham_LWSP_WaterSystem),
       subtitle = paste("Ownership:", Durham_LWSP_Ownership),
       y="Withdrawal (mgd)",
       x="Date") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
#Construct the scraping web address
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid='
the_pwsid <- '03-32-010'
the_year <- 2015
the_scrape_url <- paste0(the_base_url, the_pwsid, '&year=', the_year)
print(the_scrape_url)

#Retrieve the website contents 
the_website <- read_html(the_scrape_url)

#Create our scraping function
scrape.it <- function(the_year, the_pwsid){
  
  #Retrieve the website contents 
  the_website <- read_html(the_scrape_url)
  
  #Set the element address variables
  the_WaterSystem_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  the_PWSID_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  the_Ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  the_MaxWithdraws_tag <- 'th~ td+ td'
  
  #Scrape the data items
  the_WaterSystem <- the_website %>% html_nodes(the_WaterSystem_tag) %>% html_text()
  the_PWSID <- the_website %>% html_nodes(the_PWSID_tag) %>%  html_text()
  the_Ownership <- the_website %>% html_nodes(the_Ownership_tag) %>% html_text()
  the_MaxWithdraws <- the_website %>% html_nodes(the_MaxWithdraws_tag) %>% html_text()
  
  #Convert to a dataframe
  df_withdrawals <- data.frame("Year" = rep(the_year,12)) %>% 
    mutate(WaterSystem = !!the_WaterSystem,
           PWSID = !!the_PWSID,
           Ownership = !!the_Ownership,
           MaxDayUse = as.numeric(the_MaxWithdraws),
           Date = my(paste(LWSP_Month_numeric,"-",Year)))  
  
  #Return the dataframe
  return(df_withdrawals)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
Durham_2015 <- scrape.it(2015, '03-32-010')
view(Durham_2015)

ggplot(Durham_2015, aes(x = Date, y = MaxDayUse)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste(the_year, "Maximum Daily Withdrawals for", the_WaterSystem),
       subtitle = paste("Ownership:", the_Ownership),
       y="Withdrawal (mgd)",
       x="Date") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid='
the_pwsid <- '01-11-010'
the_year <- 2015
the_scrape_url <- paste0(the_base_url, the_pwsid, '&year=', the_year)
print(the_scrape_url)

Asheville_2015 <- scrape.it(2015, '01-11-010')
view(Asheville_2015)

Comparison_2015 <- rbind(Asheville_2015, Durham_2015)

ggplot(Comparison_2015, aes(x = Date, y = MaxDayUse, color = WaterSystem)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Maximum Daily Withdrawals for Durham and Asheville"),
       y="Withdrawal (mgd)",
       x="Date") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
#Set the inputs to scrape years 2015 to 2020 for the site "0004-0001"
the_years = rep(2010:2021)
my_pwsid = '01-11-010'

Asheville_dfs <- lapply(X = the_years,
                  FUN = scrape.it,
                  the_pwsid = my_pwsid)

Asheville_dfs <- map2(the_years, my_pwsid, scrape.it) 

#OR, use purrr's map function
Asheville_dfs <- map(the_years, scrape.it, the_pwsid = my_pwsid)

#Conflate the returned dataframes into a single dataframe
Asheville_df <- bind_rows(Asheville_dfs)
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
