---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Siyu Dong"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2, message = FALSE, error = FALSE, warning = FALSE, echo = TRUE}
#1
# Working directory checked with getwd()
library(tidyverse)
library(agricolae)
library(here)
library(ggplot2)
library(htmltools)
library(dplyr)

Lake.Chem.Raw <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
                          stringsAsFactors = TRUE)

Lake.Chem.Raw$sampledate <- as.Date(Lake.Chem.Raw$sampledate, format = "%m/%d/%y")

#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:

 * H0: There is no significant relationship between mean lake temperature recorded during July and depth across all lakes.
 * Ha: There is a significant relationship between mean lake temperature recorded during July and depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 째C. Make this plot look pretty and easy to read.

```{r scatterplot, message=FALSE, warning=FALSE}
#4
Lake.Chem.July <- Lake.Chem.Raw %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5
Depth_Temp <- ggplot(Lake.Chem.July, aes(x = depth, y = temperature_C)) +
  geom_point(color = "blue", alpha = 0.6, size = 3) +  
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  scale_y_continuous(limits = c(0, 35)) + 
  labs(x = "Depth", y = "Temperature (째C)", 
       title = "Relationship between Depth and Temperature") + 
  mytheme
print(Depth_Temp)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that the temperature decreases as the depth of the lake increases, especially from 0 to 10 meters of the lakes' depth.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
Depth_Temp_GLM <- lm(temperature_C ~ depth, data = Lake.Chem.July)
summary(Depth_Temp_GLM)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:

 * The multiple R-squared value is 0.7387, suggesting that approximately 73.87% of the variability in temperature can be explained by changes in depth.
 * The model is based on 9726 degrees of freedom, which accounts for the sample size and the number of predictor variables.
 * The p-value for the coefficient of depth is < 2.2e-16, which is extremely small. This indicates strong evidence against the null hypothesis, suggesting that there is a significant linear relationship between depth and temperature.
 * The coefficient for the depth predictor variable is -1.94621. This indicates that for every 1 meter increase in depth, the temperature decreases by approximately 1.95째C. The intercept is 21.96, representing the estimated temperature when the depth is zero (which may not be practically meaningful in this context).
Simply put, the linear regression model suggests a strong and significant negative relationship between depth and temperature in the lakes. 
---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TPAIC <- lm(data = Lake.Chem.July, temperature_C ~ year4 + daynum + depth)
step(TPAIC)
# According to the outcome, none of the predictors should be deleted and the formula goes like: 
# lm(formula = temperature_C ~ year4 + daynum + depth, data = Lake.Chem.July)

#10
TP_Temp <- lm(data = Lake.Chem.July, temperature_C ~ year4 + daynum + depth)
summary(TP_Temp)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 

 * The final set of explanatory variables suggested by the AIC method to predict temperature in the multiple regression model includes year4, daynum, and depth.
 * This model explains approximately 74.12% of the observed variance in temperature_C, as indicated by the adjusted R-squared value.
 * According to the standard error, degrees of freedom, and R-squared, this model has a slight improvement over the model using only depth as the explanatory variable.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# ANOVA
LakeTemp_ANOVA <- aov(data = Lake.Chem.July, temperature_C ~ lakename)
summary(LakeTemp_ANOVA)

# Linear Model
LakeTemp_Linear <- lm(data = Lake.Chem.July, temperature_C ~ lakename)
summary(LakeTemp_Linear)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 

 * there is a significant difference in mean temperature among the lakes. Both the ANOVA and linear regression models indicate a significant effect of the lakename variable on temperature.
 * The ANOVA test yields an F-statistic of 50 with a p-value less than 2.2e-16, indicating a significant effect of lake name on temperature.
 * In the linear model, each lake name coefficient has a significant p-value (less than 0.05), indicating that each lake has a significantly different temperature compared to the reference lake (Intercept). Additionally, the overall model has a significant F-statistic with a p-value less than 2.2e-16.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, message=FALSE, warning=FALSE}
#14.
Depth_Temp_Plot <- ggplot(Lake.Chem.July, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(method = "lm", se = FALSE, size = 1.2) +
  ylim(0, 35) + 
  labs(x = "Depth", y = "Temperature (째C)", color = "Lake Name",
       title = "Lakes' Temperature by Depth") + 
  mytheme

print(Depth_Temp_Plot)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(LakeTemp_ANOVA)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
 * Lakes with similar mean temperatures to Peter Lake: Crampton Lake, Hummingbird Lake, Ward Lake
 * Lake with statistically distinct mean temperature from all other lakes: East Long Lake
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
Two-sample t-test. The null hypothesis for the two-sample t-test would be that there is no difference in mean temperatures between Peter Lake and Paul Lake, while the alternative hypothesis would be that there is a significant difference in mean temperatures between the two lakes. If the p-value from the two-sample t-test is less than the chosen significance level (usually 0.05), we would reject the null hypothesis and conclude that there is a statistically significant difference in mean temperatures between Peter Lake and Paul Lake. Otherwise, if the p-value is greater than 0.05, we would fail to reject the null hypothesis, indicating that there is not enough evidence to conclude a significant difference in mean temperatures between the two lakes.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
July_Crampton_Ward <- subset(Lake.Chem.July, lakename %in% c("Crampton Lake", "Ward Lake"))

t_test_result <- t.test(temperature_C ~ lakename, data = July_Crampton_Ward)

print(t_test_result)
```

>Answer: 
Since the p-value (0.2649) is greater than the commonly used significance level of 0.05, we fail to reject the null hypothesis. Therefore, based on this test, we do not have sufficient evidence to conclude that there is a statistically significant difference in mean temperatures between Crampton Lake and Ward Lake in July, which does match the answer for Q16.