---
title: "Assignment 2: Coding Basics"
author: "Siyu Dong"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.
```{r}
Three_step_sequence <- seq(1,30,3)
Three_step_sequence
```

2.  Compute the mean and median of this sequence.
```{r}
mean_values <- mean(Three_step_sequence)
median_values <- median(Three_step_sequence)

mean_values
median_values
```

3.  Ask R to determine whether the mean is greater than the median.
```{r}
mean <- mean_values
median <- median_values

compare_mean_and_median <- function(mean, median){
  if (mean > median) {
    print("The mean is greater than the median.")
  }
  else if (mean < median) {
    print("The mean is smaller than the median.")
  }
  else {
    print("The mean equals to the median.")
  }
}

compare_mean_and_median(mean, median)
```


4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. I'm generating a sequence starting from 1 and ending at 30 with a interval of 3. 
# The final outcome was give the name as "Three_step_sequence". Details are as below:
Three_step_sequence <- seq(1,30,3) # Name the assigned sequence
Three_step_sequence # Print this sequence

#2. I'm trying to calculate the mean and median of the computed sequence.
mean_values <- mean(Three_step_sequence) 
# Assign the computed sequence's mean to the variable "mean_values"
median_values <- median(Three_step_sequence) 
# Assign the computed sequence's median to the variable "median_values"

mean_values # Print the result
median_values # Print the result

#3. Do the comparison between the values of the mean and the median computed from the Three_step_sequence
mean <- mean_values 
# Assign the previously computed mean value to a new variable
median <- median_values 
# Assign the previously computed median value to a new variable

compare_mean_and_median <- function(mean, median){ 
# Create the if/else function expression and divide them into 3 scenarios to discuss
  if (mean > median) {
    print("The mean is greater than the median.")
  }
  else if (mean < median) {
    print("The mean is smaller than the median.")
  }
  else {
    print("The mean equals to the median.")
  }
}

compare_mean_and_median(mean, median) 
# Based on the algorithm, compare the mean and the median and then print the result

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.
```{r}
student_name <- c("May", "Joshua", "Bob", "Smantha")
test_scores <- c(99, 61, 45, 80)
pass_or_fail <- test_scores >= 50
```


6.  Label each vector with a comment on what type of vector it is.
```{r}
class(student_name)
class(test_scores)
class(pass_or_fail)

student_name # character
test_scores # numeric
pass_or_fail #logical
```


7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r}
student_test_info <- data.frame(student_name, test_scores, pass_or_fail)
student_test_info
```


8.  Label the columns of your data frame with informative titles.

```{r basics 2}
colnames(student_test_info) <- c("Name", "Score", "Status")
student_test_info
```


9.  QUESTION: How is this data frame different from a matrix?

> Answer:
1. For data frame, there can be different data types among columns; for matrix, all the elements must maintain the same data type.
2. There could be column and row names respectively for a data frame, while no name can be assigned for matrix.


10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r}
#Option 1
check_scores <- function(student_scores){
  for (score in student_scores) {
    if (score >= 50) {
      print("True")
    }
    else if (score < 50) {
      print("False")
    }
  }
}

# Option 2
check_pass_or_fail <- function(student_scores){
  for (score in student_scores) {
    ifelse(score >= 50,
           print("True"),
           print("False"))
  }
}
  
```



11. Apply your function to the vector with test scores that you created in number 5.
```{r basics 3}
#Option 1
check_scores <- function(student_scores){
  for (score in student_scores) {
    if (score >= 50) {
      print("True")
    }
    else if (score < 50) {
      print("False")
    }
  }
}

student_scores <- student_test_info$Score

check_scores(student_scores)

# Option 2
check_pass_or_fail <- function(student_scores){
  for (score in student_scores) {
    ifelse(score >= 50,
           print("True"),
           print("False"))
  }
}

student_scores <- student_test_info$Score

check_pass_or_fail(student_scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
They both worked!